---
title: "J1790. CIBERSORT on PDAC Atlas"
author: "Joe and Luda"
date: "03/02/2022"
output: 
  html_document:
    toc: yes
    toc_float: yes
---
# Analysis plan
CIBERSORT:

* Paired Wilcoxon test for pre vs. C4D1 for only the samples we have available
* Wilcoxon test for pre only: Responders vs. nonresponders (PD + NE) (no arm stratification)
* Wilcoxon test for pre only: Time on study <90 vs. >90d (no arm stratification)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE,
                      warning = FALSE)
rm(list = ls())

library(dplyr)
library(ggfortify)
library(DT)
library(ComplexHeatmap)
library(ggpubr)

set.seed(12345)

rootFolder = "C:/Users/Luda/OneDrive - Johns Hopkins/J1790/"
outputFolder = paste0(rootFolder,"Results/Cibersort/")

timeCol = c(PreTX = 'blue',OnTX = 'red')
respCol = c(PD = 'red',SD = 'yellow')
pathCol <- c("non-significant" = "grey", "significant" = "red")
binaryCol = c('TRUE' = 'black', 'FALSE' = 'grey')

```

```{r, fig.width=10}
# Load Cibersort data from PDAC atlas
ciber <- read.csv(file = paste0(rootFolder,"Data/Atlas_Cibersort_100perm/CIBERSORTx_Job35_Results.csv"), row.names = 1)

# Load in SampleAnnot
SampAnnot <- readRDS(paste0(rootFolder,"Data/SampAnnot.rds"))

# keep cell types only (remove p-values and correlations)
ciber = ciber[rownames(SampAnnot), 1:19]

# remove cells with all zeros
ciber = ciber[,colnames(ciber)[which(!apply(ciber == 0, 2, all))]]

# top annotation of the heatmap
HA <- HeatmapAnnotation(df = SampAnnot[rownames(ciber),c("Response","timeOnStudy90", "Timepoint", "subject.ID")],
                        col=list(timeOnStudy90 = binaryCol, Response = respCol, Timepoint = timeCol))

Heatmap(t(ciber),
        top_annotation = HA,
        clustering_distance_rows = 'pearson', 
        clustering_distance_columns = 'pearson',
        row_names_gp = gpar(fontsize = 6), 
        column_title="CIBERSORT")

# correlation heatmap
cor_df <- round(cor(ciber), 2)

pheatmap(cor_df)

```


# Paired analysis for pre- vs. on-treatment

```{r, fig.width= 7, fig.height=5}
# select paired samples
pairedPatID = SampAnnot[which(duplicated(SampAnnot$subject.ID)), "subject.ID"]
pairedSamp = SampAnnot %>% filter(subject.ID %in% pairedPatID) %>% select(Sample.Name) %>% unlist()

# annotation for paired samples only
d = SampAnnot[pairedSamp, ]
pairedPre = d%>% filter(Timepoint == 'PreTX') %>% rownames()
pairedOn = d%>% filter(Timepoint == 'OnTX') %>% rownames()

# calculate p-values for paired samples
pairedPvals = sapply(colnames(ciber), function(x) wilcox.test(ciber[pairedPre,x],ciber[pairedOn,x], paired = T)$p.value )

print(pairedPvals[order(pairedPvals)])


```


## Pre-treatment only: Responders (SD) vs. nonresponders (PD)

```{r}
# select pre-treated samples only
pretxSamp = SampAnnot %>% filter(Timepoint == 'PreTX') %>% select(Sample.Name) %>% unlist()

# calculate p-values 
respPvals = sapply(colnames(ciber), function(x) wilcox.test(ciber[pretxSamp,x] ~ SampAnnot[pretxSamp,'Response'])$p.value )

print(respPvals[order(respPvals)])
write.csv(respPvals, file = paste0(rootFolder,"Results/Cibersort/Cibersort_pvals.csv"))

# make boxplots for significant ones
significant = names(respPvals)[which(respPvals < 0.05)]

for(k in significant)
{ 
p = ggplot(data.frame(proportion = ciber[pretxSamp,k], Response = SampAnnot[pretxSamp,'Response']), aes(x=Response, y=proportion)) + 
  geom_boxplot() + 
  labs(title=k) + 
  stat_compare_means() + 
  geom_point() + 
  theme_bw()

print(p)

}

```

## Pre-treatment only: time on study > 90

```{r}
# select pre-treated samples only without NA in timeOnStudy
pretxSamp = SampAnnot %>% filter(Timepoint == 'PreTX' & !is.na(timeOnStudy90)) %>% select(Sample.Name) %>% unlist()

# calculate p-values 
respPvals = sapply(colnames(ciber), function(x) wilcox.test(ciber[pretxSamp,x] ~ SampAnnot[pretxSamp,'timeOnStudy90'])$p.value )

print(respPvals[order(respPvals)])



```



