---
title: "J1790. Bulk RNASeq Analysis for the paper"
author: "Ludmila Danilova & Joe Tandurella"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_float: yes
---
# Analysis plan

- Paired differential analysis for pre vs. C4D1 for only the samples we have available
- Differential analysis for pre only: Responders vs. nonresponders (PD + NE) (no arm stratification)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE,
                      warning = FALSE)
rm(list = ls())

library(DESeq2)
library(xlsx)
library(dplyr)
library(ggfortify)
library(DT)
library(fgsea)
library(msigdbr)
library(ComplexHeatmap)
library('EnhancedVolcano')
library(data.table)

# function for making barplot wiht GSEA results
pathBarPlot = function(fgseaRes, signThr = 0.05, ttl = "")
{
  fgseaRes$adjPvalue <- ifelse(fgseaRes$padj <= 0.05, "significant", "non-significant")
  ggplot(fgseaRes, aes(reorder(pathway, NES), NES, fill = adjPvalue)) +
    geom_col() +
    scale_fill_manual(values = pathCol) +
    theme(axis.text.y = element_text(size=5)) +
    coord_flip() +
    labs(x="Pathway", y="Normalized Enrichment Score",  title=ttl) + 
    theme(plot.title.position = "plot", plot.title = element_text(size=8, face="bold"))
}
# make barplot of fgsea results
fgseaBarplot = function(res, 
                        adjPvalThr = 0.05, 
                        main_title = "Pathways Enrichment Score",
                        sig_only = FALSE)
{
  ## Complete  Barplot
  res$adjPvalue <- ifelse(res$padj <= adjPvalThr, "significant", "non-significant")
  cols <- c("non-significant" = "grey", "significant" = "red")
  if(!sig_only)
  {
    p = ggplot(res, aes(reorder(pathway, NES), NES, fill = adjPvalue)) +
      geom_col() +
      scale_fill_manual(values = cols) +
      theme(axis.text.y = element_text(size=5)) +
      coord_flip() +
      labs(x="Pathway", y="Normalized Enrichment Score", title=main_title) + 
      theme(plot.title.position = "plot", plot.title = element_text(size=8, face="bold"))
    print(p)
  }
  ## Sig Only  Barplot
  Sig_ORGResHALLMARK <- res[padj < adjPvalThr]
  if (nrow(Sig_ORGResHALLMARK)==0){ print("No Significant pathways with pAdj < 0.05"
  )} else {
    # prints the number of significant pathways
    cat("The number of significant pathways: ",nrow(Sig_ORGResHALLMARK))
    ggplot(Sig_ORGResHALLMARK, aes(reorder(pathway, NES), NES, fill = adjPvalue)) +
      geom_col() +
      scale_fill_manual(values = cols) +
      theme(axis.text.y = element_text(size=5)) +
      coord_flip() +
      labs(x="Pathway", y="Normalized Enrichment Score",
           title=paste0("Significant (pAdj<",adjPvalThr,")", main_title)) + 
      theme(plot.title.position = "plot", plot.title = element_text(size=8, face="bold"))
  }
  
}



set.seed(12345)

rootFolder = "C:/Users/ldanilo1/OneDrive - Johns Hopkins/J1790/"
outputFolder = paste0(rootFolder,"Results/RNAseq/")

timeCol = c(PreTX = 'blue',OnTX = 'red')
respCol = c(PD = 'red',SD = 'yellow')
pathCol <- c("non-significant" = "grey", "significant" = "red")
binaryCol = c('TRUE' = 'black', 'FALSE' = 'grey')

```

```{r}
# Load Gene counts
#exprsDat <- readRDS("~/OneDrive - Johns Hopkins/J1790/Data/GeneCounts.rds")
exprsDat <- readRDS(paste0(rootFolder,"Data/RNAseq/GeneCounts.rds"))

##Save for Cibersortx
#write.table(cbind(Genes = rownames(exprsDat), exprsDat), file = "~/OneDrive - Johns Hopkins/J1790/Data/J1790_Cibersort_GeneCounts.tsv", sep = '\t', quote = F, row.names = F)

```

# Load in SampleAnnot
```{r}
#SampAnnot <- read.xlsx("~/OneDrive - Johns Hopkins/J1790/Data/J1790 RNAseq Manifest Annotated FINAL.xlsx", sheetIndex = 1)
SampAnnot <- read.xlsx(paste0(rootFolder,"/Data/J1790_SampleAnnotation_12132022.xlsx"), sheetIndex = 1, endRow = 41, colIndex = 1:10)

rownames(SampAnnot) <- SampAnnot$Sample.Name

# output some statistics
table(SampAnnot$Timepoint)
length(unique(SampAnnot$subject.ID))

print(table(SampAnnot[,c("Timepoint","Bestresponse")], useNA = "ifany"))


# remove NE
#SampAnnot = SampAnnot %>% filter(Bestresponse != 'NE')
# Combine PD and NE
SampAnnot = SampAnnot %>% mutate(Response = factor(Bestresponse,
                         levels = c('NE','SD','PD'),
                         labels = c("PD",'SD','PD')))

# change C4D1 to OnTX
SampAnnot = SampAnnot %>% mutate(Timepoint = factor(Timepoint,
                         levels = c('C4D1','PreTX'),
                         labels = c("OnTX",'PreTX')))

# add time on study cut-offs 90d and 180d
SampAnnot$timeOnStudy90 = as.numeric(SampAnnot$Timeonstudy)> 90
SampAnnot$timeOnStudy180 = as.numeric(SampAnnot$Timeonstudy)> 180

SampAnnot$subject.ID = as.factor(SampAnnot$subject.ID)
SampAnnot$Timepoint = factor(SampAnnot$Timepoint, levels = c("PreTX","OnTX"))

datatable(SampAnnot)
#saveRDS(SampAnnot, file = paste0(rootFolder,'Data/SampAnnot.rds'))

#print(table(SampAnnot$Bestresponse))
#print(table(SampAnnot$Timepoint))
print(table(SampAnnot[,c("Timepoint","Bestresponse")]))

print(table(SampAnnot[,c("Timepoint","Response")]))

#print(summary(as.numeric(SampAnnot$Timeonstudy)))
#print(table(as.numeric(SampAnnot$Timeonstudy)> 90))
#print(table(as.numeric(SampAnnot$Timeonstudy)> 180))

print(table(SampAnnot[,c("Timepoint","timeOnStudy90")], useNA = "ifany"))
print(table(SampAnnot[,c("Timepoint","timeOnStudy180")], useNA = "ifany"))


exprsDat = exprsDat[, rownames(SampAnnot)]

# save RNAseq counts and sample annotation for further analysis
#save(exprsDat,SampAnnot, file = paste0(outputFolder,"RNAseqData.rda"))

# create DESeq object
dds <- DESeqDataSetFromMatrix(countData = exprsDat, colData = SampAnnot, design = ~Timepoint+Bestresponse)

# for plotting
vst <- vst(dds)
vstDat <- assay(vst)
vstDat <- vstDat[apply(vstDat,1,sd)>0,]


# select paired samples
pairedPatID = SampAnnot[which(duplicated(SampAnnot$subject.ID)), "subject.ID"]
print(c("Patients with paired samples",length(pairedPatID)))
pairedSamp = SampAnnot %>% filter(subject.ID %in% pairedPatID) %>% select(Sample.Name) %>% unlist()


```




# Differential expression analysis
## Paired analysis for pre- vs. on-treatment


``` {r}
# Get Pathways
hallmark_df = msigdbr(species = "Homo sapiens", category = "H")
hallmark_list = hallmark_df %>% split(x = .$gene_symbol, f = .$gs_name)

```

```{r}

dds_paired <- DESeqDataSetFromMatrix(countData = exprsDat[,pairedSamp], colData = SampAnnot[pairedSamp,], design = ~subject.ID + Timepoint)
dds_paired <- DESeq(dds_paired)

#resultsNames(dds_paired)

res_paired <- results(dds_paired, contrast = c("Timepoint","OnTX","PreTX"))
res_paired <- res_paired[!is.na(res_paired$padj),]

# save results
datatable(as.data.frame(res_paired))
write.csv(res_paired, file = paste0(outputFolder,"pairedDE_results.csv"))

# Volcano plot
EnhancedVolcano(res_paired, lab=row.names(res_paired), 
                x='log2FoldChange', y='padj', FCcutoff = 0.5,
                pCutoff = 0.05, title = "Paired DE Results by Time")

# Heatmap of DE significant genes
res_paired_DEGenes <- row.names(res_paired)[res_paired$padj < 0.05 &
                              abs(res_paired$log2FoldChange) > 0.5]
length(res_paired_DEGenes)

# top annotation of the heatmap
HA <- HeatmapAnnotation(df = SampAnnot[pairedSamp,c("Timepoint", "subject.ID","Response")],
                        col=list(Timepoint = timeCol, Response = respCol))

Heatmap(t(apply(vstDat[res_paired_DEGenes, pairedSamp],1,scale)),
        top_annotation = HA,
        clustering_distance_rows = 'pearson', 
        clustering_distance_columns = 'pearson',
        row_names_gp = gpar(fontsize = 6), 
        column_labels = SampAnnot[pairedSamp,"subject.ID"],
        column_title="Paired DE Results by Time")

#====
#Pathway analysis
#Extract Stats for Pathway Analysis
res_paired_stats <- res_paired$log2FoldChange
names(res_paired_stats) <- row.names(res_paired)
# GSEA on hallmark pathways  
set.seed(123456)
HALLMARK=fgsea(pathways=hallmark_list, stats=res_paired_stats)
# plot all pathways
#pathBarPlot(HALLMARK, ttl = "Hallmark pathways Enrichment Score from Paired Samples by Time")
fgseaBarplot(HALLMARK, adjPvalThr = 0.05, 
                        main_title = "Hallmark pathways Enrichment Score from Paired Samples by Time",
                        sig_only = TRUE)
datatable(HALLMARK)
fwrite(as.data.frame(HALLMARK), file=paste0(outputFolder,"pairedDE_Hallmark.csv"))

```

## Pre-treatment only: Responders (SD) vs. nonresponders (PD)

```{r}
# select pre-treated samples only
pretxSamp = SampAnnot %>% filter(Timepoint == 'PreTX') %>% select(Sample.Name) %>% unlist()

dds_resp <- DESeqDataSetFromMatrix(countData = exprsDat[,pretxSamp], colData = SampAnnot[pretxSamp,], design = ~Response)
dds_resp <- DESeq(dds_resp)
#resultsNames(dds_paired)

res_resp <- results(dds_resp, contrast = c("Response","SD","PD"))
res_resp <- res_resp[!is.na(res_resp$padj),]

# save results
datatable(as.data.frame(res_resp))
write.csv(res_resp, file = paste0(outputFolder,"preTx_response_DE_results.csv"))

# Volcano plot
EnhancedVolcano(res_resp, lab=row.names(res_resp), 
                x='log2FoldChange', y='padj', FCcutoff = 0.5,
                pCutoff = 0.05, title = "PreTX by response. Positive is higher in SD")

# Heatmap of DE significant genes
res_resp_DEGenes <- row.names(res_resp)[res_resp$padj < 0.05 &
                              abs(res_resp$log2FoldChange) > 0.5]
length(res_resp_DEGenes)

# top annotation of the heatmap
HA <- HeatmapAnnotation(df = data.frame(Response = SampAnnot[pretxSamp,c("Response")]),
                        col=list(Timepoint = timeCol, Response = respCol))

Heatmap(t(apply(vstDat[res_resp_DEGenes, pretxSamp],1,scale)),
        top_annotation = HA,
        clustering_distance_rows = 'pearson', 
        clustering_distance_columns = 'pearson',
        row_names_gp = gpar(fontsize = 6), 
        column_title="PreTX by response")

#====
#Pathway analysis
#Extract Stats for Pathway Analysis
res_resp_stats <- res_resp$log2FoldChange
names(res_resp_stats) <- row.names(res_resp)
# GSEA on hallmark pathways  
set.seed(123456)
HALLMARK=fgsea(pathways=hallmark_list, stats=res_resp_stats)
# plot  pathways
fgseaBarplot(HALLMARK, adjPvalThr = 0.05, 
                        main_title = "Hallmark pathways Enrichment Score from PreTX by response")
datatable(HALLMARK)
fwrite(as.data.frame(HALLMARK), file=paste0(outputFolder,"preTx_response_Hallmark.csv"))


```

## Pre-treatment only: time on study > 90

```{r}
# select pre-treated samples only without NA in timeOnStudy
pretxSamp = SampAnnot %>% filter(Timepoint == 'PreTX' & !is.na(timeOnStudy90)) %>% select(Sample.Name) %>% unlist()

dds_time <- DESeqDataSetFromMatrix(countData = exprsDat[,pretxSamp], colData = SampAnnot[pretxSamp,], design = ~timeOnStudy90)
dds_time <- DESeq(dds_time)
#resultsNames(dds_time)

res_time <- results(dds_time, name = "timeOnStudy90TRUE")
res_time <- res_time[!is.na(res_time$padj),]

# save results
datatable(as.data.frame(res_time))
write.csv(res_time, file = paste0(outputFolder,"preTx_timeOnStudy_DE_results.csv"))

# Volcano plot
EnhancedVolcano(res_time, lab=row.names(res_time), 
                x='log2FoldChange', y='padj', FCcutoff = 0.5,
                pCutoff = 0.05, title = "PreTX  by time on study. Positive is higher in > 90")

# Heatmap of DE significant genes
res_time_DEGenes <- row.names(res_time)[res_time$padj < 0.05 &
                              abs(res_time$log2FoldChange) > 0.5]
length(res_time_DEGenes)

# top annotation of the heatmap
HA <- HeatmapAnnotation(df = SampAnnot[pretxSamp,c("Response","timeOnStudy90")],
                        col=list(timeOnStudy90 = binaryCol, Response = respCol))

Heatmap(t(apply(vstDat[res_time_DEGenes, pretxSamp],1,scale)),
        top_annotation = HA,
        clustering_distance_rows = 'pearson', 
        clustering_distance_columns = 'pearson',
        row_names_gp = gpar(fontsize = 6), 
        column_title="PreTX by time on study")

#====
#Pathway analysis
#Extract Stats for Pathway Analysis
res_time_stats <- res_time$log2FoldChange
names(res_time_stats) <- row.names(res_time)
# GSEA on hallmark pathways  
set.seed(123456)
HALLMARK=fgsea(pathways=hallmark_list, stats=res_time_stats)
# plot all pathways
pathBarPlot(HALLMARK, ttl = "Hallmark pathways Enrichment Score from PreTX samples by time on study")

```

# Myeloid gene signature from Project 3

```{r}
myeloidGenes = unlist(read.csv(paste0(rootFolder,'Data/Myeloid_Genes.csv'), nrows = 1, header = F))
myeloidGenes = intersect(myeloidGenes, rownames(vstDat))

HA <- HeatmapAnnotation(df = SampAnnot[,c("Timepoint", "Response","timeOnStudy90","Arm")],
                        col=list(Timepoint = timeCol, timeOnStudy90 = binaryCol, Response = respCol))

Heatmap(t(apply(vstDat[myeloidGenes, ],1,scale)),
        top_annotation = HA,
        clustering_distance_rows = 'pearson', 
        clustering_distance_columns = 'pearson',
        row_names_gp = gpar(fontsize = 6), 
        column_title="Myeloid genes. All samples")

HA <- HeatmapAnnotation(df = SampAnnot[pretxSamp,c("Response","timeOnStudy90")],
                        col=list(timeOnStudy90 = binaryCol, Response = respCol))

Heatmap(t(apply(vstDat[myeloidGenes, pretxSamp],1,scale)),
        top_annotation = HA,
        clustering_distance_rows = 'pearson', 
        clustering_distance_columns = 'pearson',
        row_names_gp = gpar(fontsize = 6), 
        column_title="Myeloid genes. PreTX samples")


# top annotation of the heatmap
HA <- HeatmapAnnotation(df = SampAnnot[pairedSamp,c("Timepoint", "subject.ID","Response")],
                        col=list(Timepoint = timeCol, Response = respCol))

Heatmap(t(apply(vstDat[myeloidGenes, pairedSamp],1,scale)),
        top_annotation = HA,
        clustering_distance_rows = 'pearson', 
        clustering_distance_columns = 'pearson',
        row_names_gp = gpar(fontsize = 6), 
        column_labels = SampAnnot[pairedSamp,"subject.ID"],
        column_title="Myeloid genes. Paired samples",
        show_row_names  = F)

#==========
# do FDR for myeloid genes only
myeloidFDR = cbind(paired = p.adjust(res_paired[myeloidGenes,'pvalue'], method = 'BH'),
                   byResp = p.adjust(res_resp[myeloidGenes,'pvalue'], method = 'BH'),
                   byTime = p.adjust(res_time[myeloidGenes,'pvalue'], method = 'BH'))
rownames(myeloidFDR) = myeloidGenes
myeloidFDR = myeloidFDR[order(myeloidFDR[,1]),]
#sum(myeloidFDR < 0.05, na.rm = T)

# GSEA on myeloid genes
set.seed(123456)

myeloidGSEA= rbind(fgsea(pathways=list(paired = myeloidGenes), stats=res_paired_stats),
                fgsea(pathways=list(byResp = myeloidGenes), stats=res_resp_stats),
                fgsea(pathways=list(byTime = myeloidGenes), stats=res_time_stats))
print("Pathway analysis using myeloid genes from Project 3 as a set")
datatable(myeloidGSEA)
```

# Myeloid genes sets from xCell paper:

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5688663/

```{r}
library(GSA)
# read in selected pathways from xCell paper
myeloidPathways = GSA.read.gmt(paste0(rootFolder,'Data/moremyeloidgenelists.txt'))
# remove empty entries in the sets
myeloidPathways$genesets = lapply(myeloidPathways$genesets, function(x) x[which(x != "")])
names(myeloidPathways$genesets) = myeloidPathways$geneset.names
set.seed(123456)
myeloidPathwaysRes = list(paired = fgsea(myeloidPathways$genesets, stats=res_paired_stats),
                          byResp = fgsea(myeloidPathways$genesets, stats=res_resp_stats),
                          byTime = fgsea(myeloidPathways$genesets, stats=res_time_stats))

print("Pathway analysis using myeloid genes from xCell paper as a set./n Results from the paired analysis")
datatable(myeloidPathwaysRes$paired)
print("Results from the preTx by response analysis")
datatable(myeloidPathwaysRes$byResp)

fwrite(as.data.frame(myeloidPathwaysRes$paired), file=paste0(outputFolder,"paired_xCell_myeloid.csv"))
fwrite(as.data.frame(myeloidPathwaysRes$byResp), file=paste0(outputFolder,"preTx_response_xCell_myeloid.csv"))


# plot myeloid pathways
pathBarPlot(myeloidPathwaysRes$paired, ttl = "Myeloid pathways Enrichment Score from paired analysis")
pathBarPlot(myeloidPathwaysRes$byResp, ttl = "Myeloid pathways Enrichment Score from from PreTX samples by response")
# plot  pathways
fgseaBarplot(myeloidPathwaysRes$byResp, adjPvalThr = 0.05, 
                        main_title = "Myeloid pathways Enrichment Score from from PreTX samples by response")

pathBarPlot(myeloidPathwaysRes$byTime, ttl = "Myeloid pathways Enrichment Score from PreTX samples by time on study")

```




